---
title: "Shared Baselines"
description: |
  Share a Common Baseline
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
canonical_url: https://environments.rstudio.com/shared
---

The shared baseline strategy fits when administrators or R champions are responsible for creating an environment where less experienced users can easily share and re-run work. The defining characteristics of the strategy are:

1. There are not strict requirements on what can be installed, the main motivation is ease of sharing. 
2. Package availability is **tied to R installations** through site-wide libraries, and updates occur on a scheduled basis.

A naive approach to this strategy is for an admin to install packages into a
system library as users request them. Unfortunately, this approach is not a
strategy but actually the [Ticket System](./reproduce.html#ticket-system) danger
zone! Before diving into the implementation steps, we need to understand the problem with this approach.

### The BIG Risk

Imagine the following scenario:^[The scenario is hypothetical and simplified, you should not be concerned about the specific packages and dates used in the example.]

1. January 1st, an admin installs `tibble` into the system library. The package is installed, along with the package's dependencies. Everything is in a consistent state because the packages all originate from CRAN on the same date, and CRAN tests to ensure the "latest" packages all work together.

```{r fig.cap="Partial Dependency Graph for tibble"}
library(DiagrammeR)
grViz("
digraph first {
  node [shape = oval]
  rlang; cli; crayon;
  
  node [shape = box]
  tibble
  
  tibble->rlang
  tibble->cli
  tibble->crayon
}
")
```

2. February 1st, the admin receives a request to install `pkgdown`. In doing so,
`pkgdown` is installed along with its dependencies, **which include `rlang`,
`cli`, and `crayon`**. Already there is a problem, any users who relied on the
older versions of `cli`, `crayon`, and `rlang` could see their code break with
no warning. But, the problem gets worse! Even though its dependencies were
updated, `tibble` was not. The result is an inconsistent state, where some
packages come from February 1st and some from January 1st. This mixed set is
_not_ tested by CRAN, and can lead to an error for anyone using `tibble`.

```{r fig.cap="A dangerous situation, resulting from a partial upgrade"}
grViz("
digraph first {
  node [shape = oval
        style = filled
        fillcolor = grey]
  rlang; cli; crayon
  
  node [shape = box
        style = filled
        fillcolor = grey]
  pkgdown
  
  node [shape = box
        style = filled
        fillcolor = white]
  tibble
  
  tibble->rlang
  tibble->cli
  tibble->crayon
  pkgdown->rlang
  pkgdown->cli
  pkgdown->crayon
}
")
```

<aside>
White - January Versions
Grey - February Versions
</aside>

The benefit of the shared baseline approach is that everyone uses the same
installed packages. The problem is if an administrator updates packages, the
update could create an inconsistent state that breaks other users' code. The
main benefit has turned into a big risk!

> Note: This scenario can also occur for individual users who share a package library
across projects. The likelihood of conflict just increases if multiple users share a library.

How do we prevent this problem? One option would be for an administrator to
install all the packages at once. Unfortunately, this option rarely works in practice
because it is incredibly time intensive and users don't know upfront the entire
list of packages they'll need.

#### Frozen Repositories

A better option is to rely on a frozen repository. A frozen repository is a way
for organizations to always get a consistent set of packages, without having to
pre-install all the packages. As an example, you could rsync CRAN to an internal
server on January 1st and host it at `https://r-pkgs.example.com/cran/012019`.
Then, no matter when an admin installs new packages, they will always get a
consistent set of packages. The next time a version of R is released, the new
version of R can be associated with a new frozen repository, e.g.
`https://r-pkgs.example.com/cran/062019`, allowing users to access updated and new
packages while still remaining consistent. The specific steps for this approach
are documented below.

```{r fig.cap="Shared Basline Strategy"}
library(ggplot2)
library(tibble)
library(lubridate)
labels <- tribble(
  ~x, ~y, ~label, ~color,
  "04-01-2018", 0.75, "R 3.4.4 \n Repo Frozen \n 042018", "grey",
  "07-01-2018", 1.75, "R 3.5.1 \n Repo Frozen \n 072018", "darkblue",
  "10-15-2018", 1.65, "Projects gradually \n  migrate", "darkred"
)
text <- tribble(
  ~x, ~y, ~label, ~color,
  "05-01-2018", 1.15, "pkgdown 1.0 installed \n rlang 0.2.0 installed", "grey",
  "09-01-2018", 1.15, "tibble 1.4.2 installed \n rlang 0.2.0 still used", "grey",
  "08-01-2018", 2.15, "pkgdown 1.1.0 installed \n rlang 0.2.1 installed", "darkblue"
)
timeline1 <- tribble(
  ~x, ~y,
  '04-01-2018', 1,
  '10-01-2018', 1,
)
timeline2 <- tribble(
  ~x, ~y,
  '07-01-2018', 2,
  '12-01-2018', 2,
)

timeline1$x <- mdy(timeline1$x)
timeline2$x <- mdy(timeline2$x)
labels$x <- mdy(labels$x)
text$x <- mdy(text$x)

ggplot() + 
  theme_minimal() + 
  geom_label(data = labels, aes(x, y, label = label, color = color)) + 
  geom_text(data = text, aes(x, y, label = label, color = color)) + 
  geom_path(data = timeline1, aes(x,y), color = "black") + 
  geom_path(data = timeline2, aes(x,y), color = "darkblue")  + 
  scale_y_continuous(breaks = NULL, limits = c(0.5,2.5)) +
  scale_x_date(limits = c(as.Date(mdy("03-01-2018")),as.Date(mdy("12-01-2018")))) + 
  scale_color_manual(values = c("darkblue", "darkred", "black"), guide = FALSE) +
  labs(
    color = NULL,
    y = NULL,
    x = "Calendar Time"
    
  )
```

Overtime, managing these repositories can become tedious, [RStudio Package
Manager](https://rstudio.com/products/package-manager) provides an easy way to
automatically access snapshots and additionally optimizes disk space and
supports internal, non-CRAN packages.

### Implementation Steps

This strategy requires a "frozen repository", as described above. Organizations
can create frozen repositories manually, tie into
[MRAN](https://mran.microsoft.com/), or use [RStudio Package
Manager](https://docs.rstudio.com/rspm/admin/repositories.html#repo-versions). 

1. Install a [version of R](./R-installation.html). This results in a versioned system library:

```
/opt/R/3.4.4/lib/R/library
```

2. Create or edit the `Rprofile.site` file, to set the repo option for this version of R to point to a frozen repository. 

```{bash eval=FALSE, echo=TRUE}
# /opt/R/3.4.4/etc/Rprofile.site
local({
  options(repos = c(CRAN = "https://r-pkgs.example.com/cran/128"))
})
```


3. Run R as root, and install the desired baseline packages. Overtime, as requests for new packages come in, install them in the same way. Consistency is guaranteed because you are always installing from the same frozen repository.

```
sudo /opt/R/3.4.4./bin/R -e 'install.packages("ggplot2")'
```    

4. Users access packages on the server without any need to install, e.g.: `library(ggplot2)`

5. (Optionally) Disable the user option to change the repository setting and discourage package installation.

```{bash eval=FALSE, echo=TRUE}
# /etc/rstudio/rsession.conf
allow-r-cran-repos-edit=0
allow-package-installation=0
```

6. (Optionally) Allow users to install packages into their personal user libraries. The user library is still tied to the R version, and the repository is still frozen due to the `Rprofile.site` setting. In this case, users won't all have the same packages, but if they share code and then install packages, they'll get the same versions.

### Common Challenges and Resolutions

#### Desktop R Users

The implementation described for the shared baseline strategy assumes users are accessing R on a shared server, using a tool like RStudio Server (Pro). Often, teams of data scientists using R from their desktops also want easy collaboration and the benefits of uniform package versions. This result is possible by adapting the strategy. Desktop users simply need to set their repository option to use a frozen repository. If all users pick the same frozen repository, they'll get the benefits of the strategy. Desktop users can set the repository using the `Rprofile.site` mechanism, or using a wizard available in RStudio (v1.2+) `Tools -> Global Options -> Packages`. 

#### New or Updated Packages

What happens if a package is updated immediately after the shared baseline is
implemented? Or a new package is added? For example, what would happen if the
repository is frozen on April 1st, and April 5th a new package is added? In this
case, users would need to wait until the next release to pull in this
update. We recommend organizations roll out new versions of R (and new package sets)
every 4-6 months.

Luckily, this type of time delay will not impact most R users. The need to
access the latest and greatest packages is rare for the majority of R users,
especially new R users. We recommend allowing advanced R users who require this
type of "bleeding edge" access to use the [Snapshot and
Restore](./snapshot.html) strategy. If a critical security
issue arises that requires a package update, re-install the
version of R in a new directory, e.g. `/opt/R/3.4.4-patch/` and follow the
entire process, perhaps removing the old R version.

#### Internal Packages

The shared baseline strategy works with internal packages as long as those
packages are available in a frozen, CRAN-like repository. [RStudio Package
Manager](https://docs.rstudio.com/rspm/admin/quickstarts.html#quickstart-local)
makes it easy to include internal packages in repository checkpoints.

#### Docker

Docker can be used alongside the shared baseline strategy to ensure that
rebuilding a Docker image always returns the same sets of packages. Docker makes
the process easier, because it negates the need to manage a system library
shared by multiple users.

```{bash eval=FALSE, echo=TRUE}
FROM ubuntu
...
# To install packages
RUN R -e 'install.packages(..., repo = "https://r-pkgs.example.com/cran/042018")'
# Or set the repo option if users will install packages in the container
RUN echo 'options(repos = c(CRAN = "https://r-pkgs.example.com/cran/042018"))' > .Rprofile
```
